Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes

### 2025-06-25 ms/ms.Rmd
address RC1.2, RC1.3, RC2.2

  *Acknowledgements.* We thank the Norwegian Public Roads Administration for sharing data from ground-penetrating radar surveys. We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim. This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway. Large language models have been used during the drafting and editing of this manuscript, with author oversight. We maintain full responsibility for the scientific output, as per the European Commission's *Living guidelines on the responsible use of generative AI in research*.

@@

Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary. 
One of the most robust relationships is a negative correlation between depth and terrain slope [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g. @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.

@@

To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020]. 
Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].

@@

```{r map-sites, out.height="81%", echo = FALSE, fig.cap = "Study areas with land cover at Skrimfjella (a) and Ørskogfjellet (b). To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study. Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope 45° and azimuth 225°. AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
knitr::include_graphics("figures/map-sites.pdf")
```

At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig. \@ref(fig:map-sites)a).
In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project. 
The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % monzonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).

At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:map-sites)b).
In the AR5 dataset, 15.3 km^2^ (12.4 %) of the study area is classified as mire.
Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset). 
This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).

@@

We chose a spatial resolution of 10 m for our predictors, depth sampling, and modelling.
We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.

@@

The radiometric surveys were conduced with 75--80 m average flight altitude, 88–108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
Data were processed following standard procedures outlined by the International Atomic Energy Association.
Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.

@@

We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
At both sites we used manual probing as the primary method of measuring peat depth, and ground-penetrating radar (GPR) as a secondary method.
That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells
We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time [@parryEvaluatingApproachesEstimating2014].
Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
A complete description of our peat depth measurements is in Appendix \ref{sec:depth-measurements}.

@@

### Modelling approach

@@

```{r map-distribution, out.height="81%", echo = FALSE, fig.cap = "Spatial and statistical distributions of mean peat depth in 10 m cells at Skrimfjella, N = 372 (a) and Ørskogfjellet, N = 1878 (b). Terrain is visualized with a hillshade with slope 45° and azimuth 225°. Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
knitr::include_graphics("figures/map-distribution.pdf")
```

@@

```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 % prediction intervals (grey veritical lines) from spatial cross-validation. Points are transparent to show overlap. Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line. Marginal distributions (top) are shaded by quartile."}
knitr::include_graphics("figures/calibration_plots.pdf")
```

@@

Ground-based measurements also have uncertainty [@parryEvaluatingApproachesEstimating2014], but our data show that this error was smaller than the variation in peat depth over short distances (e.g. 100 m).
Independent probing and GPR measurements up to 2 m apart differed by an average of 29--42 cm (Fig. \@ref(fig:GPRwavevelocity)).
By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \@ref(fig:semivariograms)).
We also note that cell-level measurement error that is correlated with predictors (e.g. underestimated depth at high Multi-Resolution Valley Bottom Flatness), will go undetected by cross-validation and does not account for the limited accuracy of our models.
Ultimately, the effect of measurement error falls mostly outside the scope of DSM, since it can only be evaluated with higher-quality, independent data.

Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak [@mulderUseRemoteSensing2011].
Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost.
This point is recognized and reflected in the rise of DSM [@minasnyDigitalMappingPeatlands2019].

DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \@ref(fig:modelMetrics)).
That is: we gave DMK and the DSM equal footing for a fair comparison.
If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (<1 m, >1 m), they would have performed worse and the advantage of the DSM would be greater.
The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R^2^ of 0.16 and 0.19).
We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.

The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
By ignoring spatial autocorrelation in peat depth (i.e. the n in *scorpan*), we have not extracted all of the information about peat depth in the study area out of the training data.
If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
Moreover, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.

@@

Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R^2^ vs. mean absolute error or root mean square error; @wadouxIntegratedApproachEvaluation2022). 
It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious. 
More standardized reporting would help but not eliminate this consideration. 
For example, R^2^ is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution. 
We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem [@milaNearestNeighbourDistance2022].

@@

Koganti et al. [-@kogantiMappingPeatDepth2023] had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent. 
Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models. 
They accounted for more variability in peat depth (adjusted $R^2 = 0.71$) but had larger errors (RSME = 110 cm). 
Koganti et al.’s [-@kogantiMappingPeatDepth2023] linear regression models produced negative predictions, and it is unclear whether the values quoted above include these. 
If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe. 
An important difference between Koganti et al. [-@kogantiMappingPeatDepth2023] and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
Thus, the footprint of their detector was much smaller and they could capture variation in radiation at finer scales.

@@

Our models erred most for the deepest peats (Fig. \@ref(fig:calPlots)).
Where overprediction occurred, the error was smaller. 
This is not unexpected for the right-skewed distributions of peat depth, but it has management implications. 
Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
That makes the maps more suited for "red-lighting" than "green-lighting" peatland conversion, for example. 
Where it does not prohibit conversion (i.e. predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration). 
Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps’ characteristics.

@@

Radiometric data had little to no predictive value at either site (poor performance of *radiometric* configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in *all predictors* configuration compared to *terrain + DMK*).
These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a good explanation for the higher predictive value of radiometric variables at Skrimfjella [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radiometric total count was a much better predictor than potassium ground concentration.

@@

Another possible reason for the poor predictive value of the radiometric data could be that other physical parameters influencing the amount of intercepted radiation varied too much within sites.
Initial source strength, soil moisture, bulk density, and porosity all affect the amount of radiation that reaches the detector [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019].
Therefore, variation in these parameters could have masked the relationship between peat depth and radiometric data.
We chose the Ørskogfjellet site in part because it has a relatively homogeneous bedrock, which should reduce the variation in initial source strength.
Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.

We do not believe that the poor predictive value of the radiometric data in this study was caused by fully attenuated radioactivity — at least not in large part. 
The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results. 
In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution. 
More than a quarter of the peats in each our study areas were less than 60 cm deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].

@@

At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:modelMetrics)). 
Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp)-\@ref(fig:pdps)). 
We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g. via climate) to explain the observed patterns. 
For example, the increase in peat depth from 350 to 700 m.a.s.l. at Skrimfjella is opposite to the general pattern of deeper peats in lowland than upland Norway [@lyngstadBeskrivelserAvTorvmassivenheter2023]. 
Moreover, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation. 
Relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g. through kNNDM spatial cross-validation).

One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the marine limit after the last ice age. 
At Ørskogfjellet, the marine limit is about 75 meters above today’s sea level [Geological Survey of Norway, @hogaasDatabaseRegistreringAv2012], where the partial dependence plot of elevation shows a sharp increase in peat depth. 
In areas under the marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal. 
This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation [@liPeatDepthCarbon2025].
We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).

@@

Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water). 
Notably, slope was inferior to (at Ørskogfjellet) or highly correlated with (at Skrimfjella) these hydrological indices. 
Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature. 
Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs. \@ref(fig:varImp), \@ref(fig:pdps)). 
The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape. 
Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.

>>>>

  *Acknowledgements.* We thank the Norwegian Public Roads Administration for sharing data from ground-penetrating radar surveys. We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim. This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, \copyright Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), \copyright Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway. Large language models have been used to expedite coding of the analysis, with author oversight. Large language models have also been used during the drafting and editing of this manuscript, with author oversight. We maintain full responsibility for the scientific output, as per the European Commission's *Living guidelines on the responsible use of generative AI in research*.

@@

Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary. 
One of the most robust relationships is a negative correlation between depth and terrain slope [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024]. 
Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g. @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024; @pohjankukkaDigitalMappingPeat2025]. 
Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.

@@

To our knowledge, this is the first study to predict continuous peat depth from airborne radiometric data using machine learning. 
Where airborne radiometric data have previously been used to predict continuous peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020]. 
Where machine learning has previously been used with airborne radiometric data, it has been to predict peat extent or depth classes rather than continuous depth [e.g., @olearyDigitalSoilMapping2022; @pohjankukkaDigitalMappingPeat2025].

@@

```{r map-sites, out.height="81%", echo = FALSE, fig.cap = "Study areas with land cover at Skrimfjella (a) and Ørskogfjellet (b). To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study. Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope $45^\\circ$ and azimuth $225^\\circ$. AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
knitr::include_graphics("figures/map-sites.pdf")
```

At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig. \@ref(fig:map-sites)a).
In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project. 
The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % monzonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is $10.8^\circ$ (IQR $4.6-15.1^\circ$).

At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:map-sites)b).
In the AR5 dataset, 15.3 km^2^ (12.4 %) of the study area is classified as mire.
Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset). 
This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is $13.0^\circ$ (IQR $4.7-18.3^\circ$).

@@

We chose a spatial resolution of 10 m for our predictors, depth sampling, and modeling.
We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.

@@

The radiometric surveys were conducted with 75--80 m average flight altitude, 88–108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
Data were processed following standard procedures outlined by the International Atomic Energy Association.
Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.

@@

We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells.
We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time [@parryEvaluatingApproachesEstimating2014].
Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
A complete description of our peat depth measurements is in Appendix \ref{sec:depth-measurements}.

@@

### Modeling approach

@@

```{r map-distribution, out.height="81%", echo = FALSE, fig.cap = "Spatial and statistical distributions of mean peat depth in 10 m cells at Skrimfjella, N = 372 (a) and Ørskogfjellet, N = 1878 (b). Terrain is visualized with a hillshade with slope $45^\\circ$ and azimuth $225^\\circ$. Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
knitr::include_graphics("figures/map-distribution.pdf")
```

@@

```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 % prediction intervals (grey vertical lines) from spatial cross-validation. Points have transparency to show overlap. Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line. Marginal distributions (top) are shaded by quartile."}
knitr::include_graphics("figures/calibration_plots.pdf")
```

@@

Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak [@mulderUseRemoteSensing2011].
Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost [@minasnyDigitalMappingPeatlands2019].

DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \@ref(fig:modelMetrics)).
If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (<1 m, >1 m), they would have performed worse and the advantage of the DSM would be greater.
The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R^2^ of 0.16 and 0.19).
We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths.

The accuracy of our mapping could have been improved with a spatially explicit approach like regression kriging.
However, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.

@@

Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude [R^2^ vs. mean absolute error or root mean square error, @wadouxIntegratedApproachEvaluation2022]. 
It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious. 
More standardized reporting would help but not eliminate this consideration. 
For example, R^2^ is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution. 
We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem [@milaNearestNeighbourDistance2022].

@@

Koganti et al. [-@kogantiMappingPeatDepth2023] had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent. 
Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models. 
They accounted for more variability in peat depth (adjusted $R^2 = 0.71$) but had larger errors (RSME = 110 cm). 
Koganti et al.’s [-@kogantiMappingPeatDepth2023] linear regression models produced negative predictions, and it is unclear whether the values quoted above include these. 
If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe. 
An important difference between Koganti et al. [-@kogantiMappingPeatDepth2023] and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
Thus, the footprint of their detector was smaller and they captured variation in radiation at finer scales.

@@

Our models erred most for the deepest peats (Fig. \@ref(fig:calPlots)).
Where overprediction occurred, the error was smaller. 
This is not unexpected for the right-skewed distributions of peat depth, but it has management implications. 
For example, it makes the maps more suited for "red-lighting" than "green-lighting" peatland conversion (assuming depth is the determinative factor). 
Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.

@@

Radiometric data had little to no predictive value at either site (poor performance of *radiometric* configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in *all predictors* configuration compared to *terrain + DMK*).
These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023; @pohjankukkaDigitalMappingPeat2025].
The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a good explanation for the higher predictive value of radiometric variables at Skrimfjella [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radiometric total count was a much better predictor than potassium ground concentration.

@@

Another possible reason for the poor predictive value of the radiometric data could be that other physical parameters influencing the amount of intercepted radiation varied too much within sites.
Initial source strength, soil moisture, bulk density, and porosity all affect the amount of radiation that reaches the detector [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019].
Therefore, variation in these parameters could have masked the relationship between peat depth and radiometric data.
We chose the Ørskogfjellet site in part because it has a relatively homogeneous bedrock, which should reduce the variation in initial source strength.
Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.

We do not believe that the poor predictive value of the radiometric data in this study was caused primarily by full attenuation of radioactivity. 
The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results. 
In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution. 
More than a quarter of the peats in each of our study areas were less than 60 cm deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].

@@

At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:modelMetrics)).
This is consistent with the findings of Pohjankukka et al. [-@pohjankukkaDigitalMappingPeat2025], who mapped peat depth classes at 50 m resolution across Finland.

At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp)-\@ref(fig:pdps)).
We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g. via climate) to explain the observed patterns. 
For example, the increase in peat depth from 350 to 700 m.a.s.l. at Skrimfjella is opposite to the general pattern of deeper peats in lowland than upland Norway [@lyngstadBeskrivelserAvTorvmassivenheter2023]. 
Moreover, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation. 
Relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g. through kNNDM spatial cross-validation).

One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the marine limit after the last ice age. 
At Ørskogfjellet, the marine limit is about 75 meters above today’s sea level [Geological Survey of Norway, @hogaasDatabaseRegistreringAv2012], where the partial dependence plot of elevation shows a sharp increase in peat depth. 
In areas under the marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal. 
This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation [@liPeatDepthCarbon2025].
The same has been reported for Finland [@pohjankukkaDigitalMappingPeat2025]. 
We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).

@@

Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water). 
Notably, slope was inferior to (at Ørskogfjellet) or highly correlated with (at Skrimfjella) these hydrological indices. 
Mappers of peat depth should not assume that slope is the best terrain-derived predictor, despite its prevalence in the literature [@pohjankukkaDigitalMappingPeat2025]. 
Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs. \@ref(fig:varImp), \@ref(fig:pdps)). 
The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape. 
Including multiple scales of these variables allows the model to capture different hydrological mechanisms or patterns operating at different spatial scales.
